---
title: 'Reproducible Research Assignment #1'
author: "Paul Tan"
date: "October 1, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing the data
*We shall load the activity data using read.csv() function**
```{r}
loaded_activity <- read.csv("activity.csv", stringsAsFactors=FALSE)
```

Let's see our loaded activity data
```{R}
head(loaded_activity)
```

**Process/transform the data (if necessary) into a format suitable for your analysis**
```{r}
#CONVERT ALL DATES IN THE DATASET INTO ACTUAL R DATE FORMAT
loaded_activity$date <- as.POSIXct(loaded_activity$date, format="%Y-%m-%d")

#DETERMINE WEEKDAY AND DAYTYPE DATA COLUMN BASED ON THE ORIGINAL ACTIVITY DATE
loaded_activity <-data.frame(col_date=loaded_activity$date,
                             col_weekday=tolower(weekdays(loaded_activity$date)), 
                             col_daytype=ifelse(tolower(weekdays(loaded_activity$date)) == "saturday" | 
                                                tolower(weekdays(loaded_activity$date)) == "sunday",                                                          "WEEKEND", "WEEKDAY"),                
                             col_steps=loaded_activity$steps,
                             col_interval=loaded_activity$interval)

```                       

Checking our transformed and cleaned dataset
```{r}
head(loaded_activity)
```



## What is mean total number of steps taken per day?
**Calculate the total number of steps taken per day**
```{r}
#CALCULATE THE SUM OF STEPS GROUP BY ACTIVITY DATE
steps_per_day_summary <- with(loaded_activity,aggregate(loaded_activity$col_steps, 
                                              by=list(loaded_activity$col_date),                                                                            na.rm=TRUE, FUN=sum))
#RENAME THE COLUMN
names(steps_per_day_summary) <- c("activity_date", "total_steps")
head(steps_per_day_summary)
```  

Checking our summary data
```{r}
head(steps_per_day_summary)
```

**Histogram of the total number of steps taken each day**
```{r}
#LOAD THE GGPLOT LIBRARY
library(ggplot2)
#BEGIN PLOTTING THE SUMMARY PLOT
qplot(steps_per_day_summary$total_steps, binwidth=1000, xlab="total number of steps taken each day")
```  

**Mean and median number of steps taken each day**
```{r}
mean(steps_per_day_summary$total_steps)
median(steps_per_day_summary$total_steps)
```  

The required mean value is **`r mean(steps_per_day_summary$total_steps)`** while the median is **`r median(steps_per_day_summary$total_steps)`**


## What is the average daily activity pattern?

**Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**
```{r}
average_steps <- aggregate(loaded_activity$col_steps, by=list(loaded_activity$col_interval), 
                 FUN=mean, na.rm=TRUE)
#RENAME THE HEADER OF THE DATASET
names(average_steps) <- c("time_interval", "average_steps")
```

Let's look at the average_steps data
```{r}
head(average_steps)
```

Time series plot
```{r}
plot(average_steps$time_interval, 
     average_steps$average_steps, 
     type="l", 
     col="green", 
     lwd=2, 
     xlab="Time Interval", 
     ylab="Average Steps", 
     main="Average steps per interval Time-series")
```

**Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**
```{r}
#GET THE MAXIMUM NUMBER OF STEPS
maximum_avg_steps <- which(average_steps$average_steps == max(average_steps$average_steps))
maximum_interval <- average_steps[maximum_avg_steps, 1]
```

Let's look at the maximum_interval data
```{r}
head(maximum_interval)
```



## Imputing missing values

**Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**
```{r}
#CALCULATE THE TOTAL NUMBER OF ROWS WHICH CONTAIN NA DATA
total_NA_data <- sum(is.na(loaded_activity$col_steps))
```

**Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**
```{r}
Na_data_pos <- which(is.na(loaded_activity$col_steps))

#CREATE VECTOR OF MEANS WITH Na_data_pos AS THE REPETITIVE VALUE
vector_means <- rep(mean(loaded_activity$col_steps, na.rm=TRUE), times=length(Na_data_pos))
```

Let's look at the vector_means data
```{r}
head(vector_means)
```

**Create a new dataset that is equal to the original dataset but with the missing data filled in.**
```{r}
#REPLACE ALL THE NA ROWS WITH VECTORS THAT CONSIST OF MEAN VALUE ON THE ADDITIONAL COLUMN
loaded_activity[Na_data_pos, "col_steps"] <- vector_means
```
Let's look at the vector_means data
```{r}
head(loaded_activity)
```



**Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps**
```{r}
#CALCULATE TOTAL STEPS GROUP BY DAY WITH NA VALUES REMOVED
sum_data <- aggregate(loaded_activity$col_steps, by=list(loaded_activity$col_date), FUN=sum)

#RENAME THE COLUMN
names(sum_data) <- c("date", "total")

# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")
```

The required mean value is **`r mean(sum_data$total)`** while the median is **`r median(sum_data$total)`**



##Are there differences in activity patterns between weekdays and weekends?

**For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.**
```{r }
head(loaded_activity)
```

**Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**
```{r}
#LOAD THE LIBRARY
library(lattice)

#GET THE TOTAL MEAN STEPS GROUP BY DAYTYPE, WEEKDAY AND INTERVAL
average_steps_taken <- aggregate(loaded_activity$col_steps, by=list(loaded_activity$col_daytype, 
                               loaded_activity$col_weekday, loaded_activity$col_interval), mean)

#CHANGE THE COLUMN NAME
names(average_steps_taken) <- c("Type", "weekday", "interval", "average_mean")

```

**Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.**
```{r}
xyplot(average_mean ~ interval | Type, average_steps_taken, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))
```

